---
title: "Signal and Batch Correction for Mass Spectrometry (sbcms)"
author: 
- name: "Andris Jankevics"
  affiliation: Phenome Centre Birmingham, University of Birmingham
  email: a.jankevics@bham.ac.uk
package: sbcms
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2  
    number_sections: true  
    toc_float: true

vignette: >
  %\VignetteIndexEntry{sbcms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

R package sbcms (Signal & Batch Correction for Mass Spectrometry) is an 
implementation of Quality Control-Robust Spline Correction (QC-RSC)
[Kirwan et al](<https://dx.doi.org/10.1007/s00216-013-6856-7>) algorithm 
for signal drift and batch effect correction within/across a multi-batch direct
infusion mass spectrometry (DIMS) and liquid chromatography mass spectrometry 
(LCMS) datasets.

# Installation

You should have R version 3.4.1 or above and Rstudio installed to be able to
run this notebook.

Execute following commands from the R terminal.


```{r echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
install.packages("devtools")
install.packages("ggplot2")
install.packages("gridExtra")

devtools::install_github("computational-metabolomics/pmp")
devtools::install_github("computational-metabolomics/sbcms")
```

Load the required libraries into the R environment
```{r message=FALSE, warning=FALSE}
library(sbcms)
library(ggplot2)
library(reshape2)
library(gridExtra)
library(pmp)
```

# Data set

We will be using a small subset of 20 features of an DIMS data set consisting 
of 172 samples measured across 8 batches.
More detailed description of the data set is available from 
[Kirwan et al](https://www.nature.com/articles/sdata201412) and 
[MTBLS79](https://www.ebi.ac.uk/metabolights/MTBLS79).

```{r}
# DIMS data set is included in 'sbcms' package.
# (MTBLS79) [https://www.ebi.ac.uk/metabolights/MTBLS79]

data <- sbcdata$data
class <- sbcdata$class
batch <- sbcdata$batch
sample_order <- c(1:nrow(data))

# Input data frame or matrix should have features in rows and samples in columns
data <- t(data)

# Input data structure
data[1:5, 1:10]

class[1:10]
batch[1:10]
sample_order[1:10]
```

# Applying signal and batch correction 

Function QCRSC should be used to apply signal batch correction. 
Input peak table should be in format where features are in rows and samples in 
columns. 

Argument 'order' should be numeric vector containing sample injection 
order during analytical measurement and should be the same length as number of
columns in peak table. 

Argument `batch' should be numeric vector containing values of sample batch 
number. If all samples were measured in 1 batch, then all values in the 'batch'
vector should be '1'. 

Values for 'classes' should be character vector containing sample class labels.
Class label for quality control sample has to be "QC".

```{r}
corrected_data <- sbcms::QCRSC(df=data, order=sample_order, batch=batch, 
  classes=class, spar=0, minQC=4)
```

# Visualising results

Function 'sbcmsPlot' is providing visual comparison of data before and after 
correction. Fore example we can check output for features '1', '5', and '30' in 
peak matrix.
```{r message=FALSE, warning=FALSE, fig.height=6, fig.width=6}
plots <- sbcmsPlot (df=data, corrected_df=corrected_data, classes=class, 
  batch=batch, output=NULL, indexes=c(1, 5, 30))
plots
```

